The hedging hyperprior for the normal distribution: known mean, unknown precision

### 2023-03-14 text/23/normal-hedging-hyperprior-part-1.Rmd
Barely started work on normal-hedging-hyperprior-part-2

---
title: "The hedging hyperprior for the normal distribution"
source: "New"
date: "2023-02-28"
categories:
- Blog post
tags:
- Bayesian statistics
output: html_document
---

@@

To do:

+ Rewrite normals using a precision parameter
+ Change the hyperparameter from lambda to delta, kappa, tau, psi, or omega.

@@

[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf

@@

$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

Assume for simplicity that $\sigma$ is known and equal to $\sigma_F$. A more general case, with $\sigma$ unknown is described on a separate web page.

The mathematical details are a bit easier to follow if you let

$\tau_F = \frac{1}{\sigma_F^2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance, The normal density function in terms of $\tau_F$ is

$f(X; \mu, \tau_F) = \sqrt{\frac{\tau_F}{2 \pi}}e^{- \frac{1}{2}\tau_F (X-\mu)^2}$

Simplify this by removing any constants not involving $X$, $\mu$, or $\tau_F$.

$f(X) \propto e^{- \frac{1}{2} \tau_F (X-\mu)^2}$

You have an informative prior on $\mu$ that also has a normal distribution with a prior mean $\mu_0$ and a prior precision $\tau_0$.

$g(\mu; \mu_0, \tau_0) \propto  e^{- \frac{1}{2} \tau_0 (\mu-\mu_0)^2}$

@@

[Note that the following formulas need a lot of work.]

Drop the middle exponential term, which is constant with respect to $\mu$. Set

$\mu_W=\frac{\tau_0 \mu_0 + n \tau_F \bar{X}}{\tau_0 + n \tau_F}$

and rewrite the other two exponentials as

$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu - \mu_W + \mu_W - mu_0)^2} \tau_F^n e^{-\tau_F(n-1)S^2} e^{-n \tau_F^2(\mu - \mu_W + \mu_W - \bar{X})^2)}$

which simplifies to

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)^2(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

which, thanks to the miracle of conjugancy, is a normal distribution. The posterior mean

@@

```{r}
x_max <- 10

mu0 <- 8
tau0 <- 4

tauF <- 10

n <- 10
xbar <- 5

@@

prior05 <- qnorm(0.025, mu0, tau0) 
prior95 <- qnorm(0.975, mu0, tau0)

post05 <- qnorm(0.025, mu1, tau1) 
post95 <- qnorm(0.025, mu1, tau1)

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r sprintf("%.2f", mu0)` and it excludes `r xbar`, the mean of the data.

@@

Define your prior distribution as an inverse gamma distributions with parameters $1+\lambda (\alpha-1)$ and $\lambda \beta$ where $\lambda$ ranges between 0 and 1. Then place a hyperprior on the parameter $\lambda$. The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.

@@

@@

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[11:40, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="mu",
  ylab="lambda")
```

@@

>>>>

---
title: "The hedging hyperprior for the normal distribution: unknown mean, known precision"
source: "New"
date: "2023-02-28"
categories:
- Blog post
tags:
- Bayesian statistics
output: html_document
---

@@

@@

@@

For this derivation, it is easier to write the density in terms of

$f(X; \mu, \tau) = \sqrt{\frac{\tau}{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

where

$\tau=\frac{1}{\sigma*2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,

Assume that $\tau$ is known and equal to $\tau_F$. A more general case, with $\tau$ unknown is described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\mu$

$f(X; \mu) \propto e^{-\frac{1}{2}\tau_F(X-\mu)^2}$

The conjugate prior for a normal distribution is also a normal distribution.

$g(\mu; \mu_0, \tau_0) \propto e^{- \frac{1}{2}\tau_0(\mu-\mu_0)^2}$

where the prior mean is $\mu_0$ and the prior precision is $\tau_0$.

@@

Drop the middle exponential term, which is constant with respect to $\mu$ and use high school algebra ([completing the square][wik1]) to get

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

which is a normal distribution. The posterior mean

@@

$\frac{n \tau_F}{n\tau_F+\tau_0}\bar{X} + \frac{\tau_0}{n\tau_F+\tau_0}\mu_0$

```{r}
x_max <- 8

mu0 <- 4
tau0 <- 8

tauF <- 0.6
n <- 10
xbar <- 2

@@

prior05 <- qnorm(0.025, mu0, 1/sqrt(tau0)) 
prior95 <- qnorm(0.975, mu0, 1/sqrt(tau0))

post05 <- qnorm(0.025, mu1, 1/sqrt(tau1)) 
post95 <- qnorm(0.975, mu1, 1/sqrt(tau1))

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.

@@

Define your prior distribution as a normal distributions with prior mean $\mu0$ and prior precision

$\tau_F+\lambda (\tau_0-\tau_F)$

where $\lambda$ ranges between 0 and 1 and $\tau_F$ is less than $\tau_0$. When $\lambda=0$, the prior precision reduces to $\tau_F$ which is equivalent to a prior sample size of $n_0=1$. When $lambda=1$, the prior precision is $\tau_0$ which represents the full strength of the informative priori distribution.

Place a hyperprior on the parameter $\lambda$. The uniform(0,1) distribution is a reasonable choice for the hyperprior, but others are possible.

@@

This is the joint prior distribution of $\mu_0$ and $\lambda$.

@@

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  mu_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(50, 75, 90)/100,
  xlab="mu",
  ylab="lambda")
```

@@

[jor1]: https://people.eecs.berkeley.edu/~jordan/courses/260-spring10/lectures/lecture5.pdf
[mur1]: https://www.cs.ubc.ca/~murphyk/Papers/bayesGauss.pdf
[wik1]: https://en.wikipedia.org/wiki/Completing_the_square

### 2023-03-14 text/23/normal-hedging-hyperprior-part-2.Rmd
Barely started work on normal-hedging-hyperprior-part-2

---
title: "The hedging hyperprior for the normal distribution: unknown mean, known precision"
source: "New"
date: "2023-02-28"
categories:
- Blog post
tags:
- Bayesian statistics
output: html_document
---

>>>>

---
title: "The hedging hyperprior for the normal distribution: known mean, unknown precision"
source: "New"
date: "2023-02-28"
categories:
- Blog post
tags:
- Bayesian statistics
output: html_document
---
